A comps-based approach for interpreting tree-based predictions with an application to the NFL draft

### 2025-08-01 articles/jsa/JSASubmission.tex
upload JSA submission files

\documentclass{article}

\usepackage[alldates=year, block=space, style=apa]{biblatex}
\setlength\bibitemsep{0.5\baselineskip}
\addbibresource{../random-forest-comps.bib}

\usepackage[margin=1in]{geometry}
\linespread{1.2}

@@

\usepackage{amsmath, amsfonts, array, authblk, float, graphicx, subcaption}

\title{A comps-based approach for interpreting tree-based predictions with an application to the NFL draft}
\author[1]{Elisabeth Millington}
\author[2]{Scott Powers}
\affil[1]{Department of Kinesiology, Rice University}
\affil[2]{Department of Sport Management, Rice University}
\date{August 1, 2025}

@@

@@

\begin{abstract}
  Random forests are a powerful yet opaque machine learning predictive model. This poses challenges for interpretability, particularly in environments where decisions can be very high-stakes, like professional sports. In this paper, we present a method for interpreting random forest predictions in a sports setting through the lens of player comps. Player comps are a common practice in player scouting, where prospects are evaluated based on their perceived similarity to past players. We can define interpretable similarity scores that quantify how much each training observation contributes to a given prediction by leveraging the connection between random forests and adaptive nearest neighbors. A player's similarity scores can be viewed as quantifiable player comps, and we show that the random forest's prediction can be recovered as the weighted average of the outcome variable among those comps, using the similarity scores as weights. We apply this methodology to evaluate quarterback prospects in the 2025 National Football League Draft using ESPN's Total Quarterback Rating, averaged across seasons. We show that our approach captures meaningful structure in the data while providing interpretability, allowing us to identify comps for top prospects such as Cam Ward, Dillon Gabriel, Shedeur Sanders, and Jaxson Dart.
\end{abstract}

@@

Random forests are a powerful machine learning prediction model that use the ensemble learning technique of combining multiple decision trees to mitigate overfitting. Each tree is built from a random bootstrap of the data, and the final prediction is determined by either majority tree vote (classification tasks) or averaging the tree responses (regression tasks). Random forests have the flexibility to model many different functional relationships while maintaining robustness to noise and outliers within the data (\cite{breiman_random_2001}), which makes them suitable for many real-world applications.

Like other so-called ``black box'' machine learning algorithms, one drawback of the random forest is that the reasons for an individual's prediction can be difficult to interpret. Interpretability is crucial when using a machine learning algorithm because it allows the user to choose the best prediction method and it builds trust for the user in the prediction (\cite{ribeiro_why_2016}). This is especially important in sports contexts because executives need to consider non-quantifiable information in their decision-making. The better a user understands how a model arrives at a prediction, the better they can determine how much it is appropriate to adjust the prediction based on subjective information from outside of the model.

Perhaps one of the most important decisions that sports executives make is selecting quarterbacks in the National Football League (NFL) draft. Quarterbacks are the backbone of offensive schemes for every NFL team and are arguably the most important position (\cite{hughes_positional_2015}). Due to the nature of the position, trading for or signing a top quarterback during free agency is uncommon, which is why drafting a quarterback with high NFL potential is crucial for team success. Since quarterbacks are highly sought after, talented college quarterbacks with perceived NFL starting potential are typically taken in the first or second round of the draft. In the NFL, this can have significant consequences because the rookie salary scale means large financial commitments to players taken at the top of the draft. In 2025, the slotted contract for No. 1 pick Cam Ward was \$48.4 million over four years, of which \$32.2 million was guaranteed as signing bonus (\cite{badenhausen_nfl_2025}).

In traditional scouting for sports, draft prospect discussions and predictions are often based on comparing a current prospect with similar draft prospects in the past. Some of these past draft prospects will have gone on to successful careers, and others will have experienced less success (\cite{trapasso_nfl_2025}). Scouts, managers, and coaches can assess NFL potential by comparing prospects to past players with similar collegiate careers. These historical reference players are known colloquially as ``comps'' (\cite{jones_nfl_2025}), short for comparables. The simplest example of this line of reasoning is, ``This prospect reminds me most of Player X, and Player X had a successful career, so I think this player will have a successful career,'' effectively a human implementation of the $k$-nearest neighbor ($k$-NN) algorithm, with $k = 1$.

Random forests can be thought of as an adaptive $k$-NN algorithm (\cite{lin_random_2006}), with the nearness of neighbors adaptively determined to minimize prediction error. Our work is motivated by this connection between random forests and the traditional lens of player comps. In this paper, we develop methods and software for extracting similarity scores from fitted random forest models such that the prediction for a new data point is the weighted average of the outcomes in the training set, weighted by these similarity scores.

@@

Previous research has attempted to introduce interpretability to predictions from random forests and other tree-based machine learning algorithms. \textcite{petkovic_improving_2018} describe a user-centered approach to reporting random forest results in an interpretable manner. \textcite{aria_comparison_2021} compare two different techniques for interpreting the interactions between predictor variables and response variables in random forest models. In a sports context, \textcite{ouyang_integration_2024} use Shapley Additive exPlanations (SHAP) values to interpret NBA game outcome predictions from an XGBoost model.

Draft models are common in professional sports, as evaluating prospect talent can be key to success. One example of a draft model that uses ESPN's Total Quarterback Rating as the response variable is \textcite{craig_predicting_2021}. They used logistic regression to estimate whether a quarterback would be selected in the draft and used a second regression to estimate that quarterback's NFL performance based on their final season metrics, scout grades, and player height. Their predictions are easy to interpret directly from the logistic regression model. \textcite{wolfson_quarterback_2011} use their own metric of net points combined with games played to define NFL success. They estimate a logistic regression model for this metric on a combination of NFL Combine and college statistics, and they use this model to analyze how effectively NFL teams leverage data available when drafting quarterbacks. Again, the logistic regression model results in predictions that are easily interpretable. \textcite{berger_jumping_2021} use principal component analysis (PCA) combined with a high-order regression model to evaluate NBA prospects using their college statistics and combine data. The interpretation of the principal components is unclear, resulting in a relatively uninterpretable draft model. \textcite{luo_improving_2024} propose a method for improving the way that NHL teams draft players by using both scouting reports and statistics. They use a combination of large language models and ensemble machine learning, which are both opaque in the way that they generate predictions, leading to uninterpretable predictions/reports.

@@

All data used for this study are publicly available from Sports Reference (\cite{sports_reference_sports_2025}). Our dataset includes 2,099 quarterbacks who (1) played their final season at the NCAA Division I Football Bowl Subdivision (FBS) between 2001 and 2023 (inclusive); (2) played more than six total games in their careers; and (3) attempted more than five passes per season. The data for each player-season include team, conference, strength of schedule, games played, passing statistics (attempts, completions, touchdowns, interceptions), rushing statistics (attempts, yards, touchdowns), and awards (All-America designation, Heisman voting).

In addition to college statistics, we obtained each player's passing attempts and Total Quarterback Rating (QBR) (\cite{burke_how_2016}) from each season between 2006 and 2024 (inclusive). Only available since 2006, QBR is based on expected points added (EPA), which is the change in drive scoring expectation before and after each play. It considers each quarterback's share of their team's EPA and accounts for home-field advantage, defensive strength, and garbage time. The range of QBR is from 0 to 100, and \textcite{burke_how_2016} suggests thinking of it as a percentile of performance at the individual game level. For each quarterback, we calculate their average QBR across seasons, and we define a player who has attempted fewer than 10 career NFL passes to have a QBR of zero. Here the scale of QBR is helpful because players who reach the NFL cannot be rated below players who never reach the NFL.

@@

A response variable such as nflWAR (\cite{yurko_nflwar_2019}) or Sports Reference's Approximate Value could have avoided some of the limitations of QBR, but we were unable to obtain these data for a sufficient sample of players' careers. The primary purpose of the present work (extracting similarity scores to interpret random forest predictions through the lens of player comps) holds regardless of the response variable used.

@@

Random forests are a machine learning method typically used for classification or regression (\cite{breiman_random_2001}). The algorithm creates $B$ bootstrap samples of the original data and then estimates a decision tree on each bootstrap sample, randomly choosing at each split a subset of features as candidates. The random forest's prediction is the average of the predictions of the individual trees. By fitting multiple decision trees, random forests generally achieve much lower-variance predictions than individual decision trees. The randomness introduced by the bootstrap resampling of the data and by randomly limiting eligible features at each split drives down the correlation between the trees in the forest, further reducing the variance of the fitted values.

We trained the random forest model using the R package ranger (\cite{wright_ranger_2024}), which is a fast implementation of the random forest algorithm. The response variable was average QBR across NFL seasons (zero for players who never appeared in the NFL). The college statistics used as predictor variables were games played per season, attemtped passes per season, complete passes per season, passing yards per season, passing yards per attempt, passing touchdowns per season, passing interceptions per season, rushing attempts per season, rushing yards per season, rushing touchdowns per season, number of All-America selections, final-season Hesiman Award voting finish, final-season strength of schedule, final-season games played, final-season passing yards per attempt, and final-season passer efficiency rating.

To train the random forest model, we first implemented a random 70\%/30\% train/test split. We chose $B$ = 1,000 decision trees, each trained on a bootstramp sample of the data, with replacement. To determine the number of variables to consider at each node and the maximum tree size, we used the R package tuneRanger (\cite{probst_tuneranger_2024}), which uses model-based optimization of out-of-bag prediction error to tune the hyperparameters. After 100 iterations of the tuning algorithm, we estimated the optimal number of candidate variables at each split to be 5 and the optimal minimum node size to be 73.

@@

Our method is motivated by the observation that a random forest is effectively $k$-NN, using neighborhoods adaptively determined to optimize predictive accuracy (\cite{lin_random_2006}). In what follows, $\vec x_1, ..., \vec x_n \in \mathbb{R}^p$ represent the feature vectors in the training data, and $y_1, ..., y_n \in \mathbb{R}$ represent the outcome values in the training data. We draw $B$ bootstrap samples of the data indexed by $b \in \{1, ..., B\}$, and we fit a decision tree on each of these samples. We use $n_{b,i}$ to denote the number of times observation $i$ is included in bootstrap sample $b$, and we use $\mathcal{T}_{b,i}$ to denote the terminal node of observation $i$ in decision tree $b$. We use $|\mathcal{T}_{b,i}|$ to denote the number of training observations (with repetition) which belong to $\mathcal{T}_{b,i}$. If observation $i$ is not included in bootstrap sample $b$, then $n_{b,i} = 0$, and we define $\mathcal{T}_{b,i}$ as the terminal node to which observation $i$ would be assigned, according to the rules of decision tree $b$.

From this setup, if we query the model at a new feature vector $\vec x_0 \in \mathbb{R}^p$, then the prediction of decision tree $b$ is given by
$$
  \hat{y}_b(\vec x_0) = \sum_{i = 1}^{n} w_{b,0,i} \cdot y_i,
$$
where $w_{b,0,i} = \mathbb{I}\{\mathcal{T}_{b,0} = \mathcal{T}_{b,i}\} \cdot n_{b,i} / |\mathcal{T}_{b,0}|$ is the proportion of training data in $\mathcal{T}_{b,0}$ from observation $i$. The weights $w_{b,0,\cdot}$ can be interpreted as defining a local neighborhood around $\vec x_0$ for tree $b$ (\cite{lin_random_2006}).

@@

\begin{figure}
    \centering
    \includegraphics[width=0.5\linewidth]{figures/variable_importance.pdf}
    \caption{\it Importance of predictor variables in random forest model. The response variable in this model is regressed NFL QBR, averaged across seasons in a player's career.}
    \label{fig:importance}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.5\linewidth]{figures/3d_plot.pdf}
    \caption{\it Visualization of prediction function learned by random forest model. Passing yards per season and final-season strength of schedule are two of most important predictor variables in the model. Each point represents one player from the historical training set, and the color corresponds to the player's fitted value.}
    \label{fig:3d-plot}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\linewidth]{figures/predicted_vs_actuals.pdf}
  \caption{\it Predicted and actual (regressed) QBR for all quarterbacks in the dataset. The vast majority (91\%) of actual QBRs are zero, meaning that the points along the x-axis represent 91\% of the data.}
  \label{fig:predicted-vs-actuals}
\end{figure}

@@

\begin{table}[H]
  \centering
  \input{tables/top_ten.tex}
  \caption{\it Top ten career-average QBR predictions for quarterback prospects in the 2025 draft class. This includes draft-eligible players who played at an NCAA Division I FBS school during the 2024 season.}
  \label{tab:top-ten}
\end{table}

@@

These four quarterback prospects illustrate the model's interpretability and its alignment with real-world outcomes. Sanders and Ward were both initially expected to go in the first round of the 2025 draft. Sanders' draft position unexpectedly slid, and he ended up being taken in the fifth round in the draft. However, this is likely a result of his refusal to participate in the NFL Scouting Combine, coupled with a reported bad attitude in his interviews, rather than his college performance (\cite{mckenna_what_2025}). In the end, Dart was the other quarterback taken in the first round of the draft.

\begin{table}[H]
  \resizebox{\textwidth}{!}{
    \input{tables/side_by_side_similarity.tex}
  }
  \caption{\it Top ten similarity scores for each of the top four quarterback prospects in the 2025 NFL draft. The similarity score is the contribution made by each historical prospect to the weighted average which constitutes the reference prospect's prediction in the random forest model.}
  \label{tab:side-by-side-similarity}
\end{table}

It is worth noting that both Ward and Sanders started their college careers at teams that are not in the NCAA Division I FBS, as Ward played his first season at Incarnate Word and Sanders played his first two seasons at Jackson State. As a result, their college statistics used in the model do not include their statistics from these schools. We observe that Ward has the highest QBR prediction, which intuitively makes sense because he is widely regarded as the best quarterback in this year's draft class. Gabriel has a slightly higher QBR prediction than Sanders although Sanders was considered a better prospect. However, Gabriel finished higher than Sanders in Heisman voting, and Oregon had a much better record than Colorado. Dart's relatively low prediction compared to the other three makes sense because he wasn't considered a top quarterback prospect until his performance in the NFL Combine, which this model does not take into consideration. He also performed well in his interviews with teams (\cite{traina_qb_2025}), which is not quantifiable.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.24\linewidth]{figures/prospect_histogram_ward.pdf}
    \includegraphics[width=0.24\linewidth]{figures/prospect_histogram_sanders.pdf}
    \includegraphics[width=0.24\linewidth]{figures/prospect_histogram_gabriel.pdf}
    \includegraphics[width=0.24\linewidth]{figures/prospect_histogram_dart.pdf} \\
    \includegraphics[width=0.24\linewidth]{figures/prospect_similarity_ward.pdf}
    \includegraphics[width=0.24\linewidth]{figures/prospect_similarity_sanders.pdf}
    \includegraphics[width=0.24\linewidth]{figures/prospect_similarity_gabriel.pdf}
    \includegraphics[width=0.24\linewidth]{figures/prospect_similarity_dart.pdf}
    \caption{\it Top: Histogram of training players' regressed QBR values weighted by similarity to the prospect being predicted. The vertical green line annotates each prospect's predicted QBR, which matches the mean of the weighted distribution. Bottom: Scatter plot showing individual similarity scores and regressed QBR for historical prospects.}
    \label{fig:prospect-plots}
\end{figure}

@@

\section*{Acknolwedgments}

@@

\section*{Code and Data Availability}

@@

\printbibliography

>>>>

\documentclass[Review, sageh, times]{sagej}

\usepackage{amsmath, amsfonts, array, float, graphicx, subcaption, moreverb, url, natbib}

@@

@@

\runninghead{Millington and Powers}

\title{A comps-based approach for interpreting tree-based predictions with an application to the NFL draft}

\author{Elisabeth Millington\affilnum{1} and Scott Powers\affilnum{2}}

\affiliation{\affilnum{1}Department of Kinesiology, Rice University\\
\affilnum{2}Department of Sport Management, Rice University}

\corrauth{Scott Powers, 6100 Main St, Houston, TX, 77005}

\email{scott.powers@rice.edu}

\begin{abstract}
  Random forests are a powerful yet opaque machine learning predictive model. This poses challenges for interpretability, particularly in environments where decisions can be very high-stakes, like professional sports. In this paper, we present a method for interpreting random forest predictions in a sports setting through the lens of player comps. Player comps are a common practice in player scouting, where prospects are evaluated based on their perceived similarity to past players. We can define interpretable similarity scores that quantify how much each training observation contributes to a given prediction by leveraging the connection between random forests and adaptive nearest neighbors. A player's similarity scores can be viewed as quantifiable player comps, and the random forest's prediction can be recovered by taking a weighted average of the outcome variable among those comps, with weights determined by the similarity scores. We apply this methodology to evaluate quarterback prospects in the 2025 National Football League Draft using ESPN's Total Quarterback Rating (QBR) per season. We show that our approach captures meaningful structure in the data while providing interpretability, allowing us to identify comps for top prospects such as Cam Ward, Jaxson Dart, Shedeur Sanders, and Dillon Gabriel.
\end{abstract}

\keywords{National Football League, interpretable machine learning, draft projections, player comps, random forests, sports analytics}

@@

@@

Random forests are a powerful machine learning prediction model that use the ensemble learning technique of combining multiple decision trees to mitigate overfitting. Each tree is built from a random bootstrap of the data, and the final prediction is determined by either majority tree vote (classification tasks) or averaging the tree responses (regression tasks). Random forests have the flexibility to model many different functional relationships while maintaining robustness to noise and outliers within the data \citep{breiman_random_2001}, which makes them suitable for many real-world applications.

Like other so-called ``black box'' machine learning algorithms, one drawback of the random forest is that the reasons for an individual's prediction can be difficult to interpret. Interpretability is crucial when using a machine learning algorithm because it allows the user to choose the best prediction method and it builds trust for the user in the prediction \citep{ribeiro_why_2016}. This is especially important in sports contexts because executives need to consider non-quantifiable information in their decision-making. The better a user understands how a model arrives at a prediction, the better they can determine how much it is appropriate to adjust the prediction based on subjective information from outside of the model.

Perhaps one of the most important decisions that sports executives make is selecting quarterbacks in the National Football League (NFL) draft. Quarterbacks are the backbone of offensive schemes for every NFL team and are arguably the most important position \citep{hughes_positional_2015}. Due to the nature of the position, trading for or signing a top quarterback during free agency is uncommon, which is why drafting a quarterback with high NFL potential is crucial for team success. Since quarterbacks are highly sought after, talented college quarterbacks with perceived NFL starting potential are typically taken in the first or second round of the draft. In the NFL, this can have significant consequences because the rookie salary scale means large financial commitments to players taken at the top of the draft. In 2025, the slotted contract for No. 1 pick Cam Ward was \$48.4 million over four years, of which \$32.2 million was guaranteed as signing bonus \citep{badenhausen_nfl_2025}.

In traditional scouting for sports, draft prospect discussions and predictions are often based on comparing a current prospect with similar draft prospects in the past. Some of these past draft prospects will have gone on to successful careers, and others will have experienced less success \citep{trapasso_nfl_2025}. Scouts, managers, and coaches can assess NFL potential by comparing prospects to past players with similar collegiate careers. These historical reference players are known colloquially as ``comps'' \citep{jones_nfl_2025}, short for comparables. The simplest example of this line of reasoning is, ``This prospect reminds me most of Player X, and Player X had a successful career, so I think this player will have a successful career,'' effectively a human implementation of the $k$-nearest neighbor ($k$-NN) algorithm, with $k = 1$.

Random forests can be thought of as an adaptive $k$-NN algorithm \citep{lin_random_2006}, with the nearness of neighbors adaptively determined to minimize prediction error. Our work is motivated by this connection between random forests and the traditional lens of player comps. In this paper, we develop methods and software for extracting similarity scores from fitted random forest models such that the prediction for a new data point is the weighted average of the outcomes in the training set, weighted by these similarity scores.

@@

Previous research has attempted to introduce interpretability to predictions from random forests and other tree-based machine learning algorithms. \citet{petkovic_improving_2018} describe a user-centered approach to reporting random forest results in an interpretable manner. \citet{aria_comparison_2021} compare two different techniques for interpreting the interactions between predictor variables and response variables in random forest models. In a sports context, \citet{ouyang_integration_2024} use Shapley Additive exPlanations (SHAP) values to interpret NBA game outcome predictions from an XGBoost model.

Draft models are common in professional sports, as evaluating prospect talent can be key to success. One example of a draft model that uses ESPN's Total Quarterback Rating as the response variable is \citet{craig_predicting_2021}. They used logistic regression to estimate whether a quarterback would be selected in the draft and used a second regression to estimate that quarterback's NFL performance based on their final season metrics, scout grades, and player height. Their predictions are easy to interpret directly from the logistic regression model. \citet{wolfson_quarterback_2011} use their own metric of net points combined with games played to define NFL success. They estimate a logistic regression model for this metric on a combination of NFL Combine and college statistics, and they use this model to analyze how effectively NFL teams leverage data available when drafting quarterbacks. Again, the logistic regression model results in predictions that are easily interpretable. \citet{berger_jumping_2021} use principal component analysis (PCA) combined with a high-order regression model to evaluate NBA prospects using their college statistics and combine data. The interpretation of the principal components is unclear, resulting in a relatively uninterpretable draft model. \citet{luo_improving_2024} propose a method for improving the way that NHL teams draft players by using both scouting reports and statistics. They use a combination of large language models and ensemble machine learning, which are both opaque in the way that they generate predictions, leading to uninterpretable predictions/reports.

@@

All data used for this study are publicly available from Sports Reference \citep{sports_reference_sports_2025}. Our dataset includes 2,099 quarterbacks who (1) played their final season at the NCAA Division I Football Bowl Subdivision (FBS) between 2001 and 2023 (inclusive); (2) played more than six total games in their careers; and (3) attempted more than five passes per season. The data for each player-season include team, conference, strength of schedule, games played, passing statistics (attempts, completions, touchdowns, interceptions), rushing statistics (attempts, yards, touchdowns), and awards (All-America designation, Heisman voting).

In addition to college statistics, we obtained each player's passing attempts and Total Quarterback Rating (QBR) \citep{burke_how_2016} from each season between 2006 and 2024 (inclusive). Only available since 2006, QBR is based on expected points added (EPA), which is the change in drive scoring expectation before and after each play. It considers each quarterback's share of their team's EPA and accounts for home-field advantage, defensive strength, and garbage time. The range of QBR is from 0 to 100, and \citet{burke_how_2016} suggests thinking of it as a percentile of performance at the individual game level. For each quarterback, we calculate their average QBR across seasons, and we define a player who has attempted fewer than 10 career NFL passes to have a QBR of zero. Here the scale of QBR is helpful because players who reach the NFL cannot be rated below players who never reach the NFL.

@@

A response variable such as nflWAR \citep{yurko_nflwar_2019} or Sports Reference's Approximate Value could have avoided some of the limitations of QBR, but we were unable to obtain these data for a sufficient sample of players' careers. The primary purpose of the present work (extracting similarity scores to interpret random forest predictions through the lens of player comps) holds regardless of the response variable used.

@@

Random forests are a machine learning method typically used for classification or regression \citep{breiman_random_2001}. The algorithm creates $B$ bootstrap samples of the original data and then estimates a decision tree on each bootstrap sample, randomly choosing at each split a subset of features as candidates. The random forest's prediction is the average of the predictions of the individual trees. By fitting multiple decision trees, random forests generally achieve much lower-variance predictions than individual decision trees. The randomness introduced by the bootstrap resampling of the data and by randomly limiting eligible features at each split drives down the correlation between the trees in the forest, further reducing the variance of the fitted values.

We trained the random forest model using the R package ranger \citep{wright_ranger_2024}, which is a fast implementation of the random forest algorithm. The response variable was average QBR across NFL seasons (zero for players who never appeared in the NFL). The college statistics used as predictor variables were games played per season, attemtped passes per season, complete passes per season, passing yards per season, passing yards per attempt, passing touchdowns per season, passing interceptions per season, rushing attempts per season, rushing yards per season, rushing touchdowns per season, number of All-America selections, final-season Hesiman Award voting finish, final-season strength of schedule, final-season games played, final-season passing yards per attempt, and final-season passer efficiency rating.

To train the random forest model, we first implemented a random 70\%/30\% train/test split. We chose $B$ = 1,000 decision trees, each trained on a bootstramp sample of the data, with replacement. To determine the number of variables to consider at each node and the maximum tree size, we used the R package tuneRanger \citep{probst_tuneranger_2024}, which uses model-based optimization of out-of-bag prediction error to tune the hyperparameters. After 100 iterations of the tuning algorithm, we estimated the optimal number of candidate variables at each split to be 5 and the optimal minimum node size to be 73.

@@

Our method is motivated by the observation that a random forest is effectively $k$-NN, using neighborhoods adaptively determined to optimize predictive accuracy \citep{lin_random_2006}. In what follows, $\vec x_1, ..., \vec x_n \in \mathbb{R}^p$ represent the feature vectors in the training data, and $y_1, ..., y_n \in \mathbb{R}$ represent the outcome values in the training data. We draw $B$ bootstrap samples of the data indexed by $b \in \{1, ..., B\}$, and we fit a decision tree on each of these samples. We use $n_{b,i}$ to denote the number of times observation $i$ is included in bootstrap sample $b$, and we use $\mathcal{T}_{b,i}$ to denote the terminal node of observation $i$ in decision tree $b$. We use $|\mathcal{T}_{b,i}|$ to denote the number of training observations (with repetition) which belong to $\mathcal{T}_{b,i}$. If observation $i$ is not included in bootstrap sample $b$, then $n_{b,i} = 0$, and we define $\mathcal{T}_{b,i}$ as the terminal node to which observation $i$ would be assigned, according to the rules of decision tree $b$.

From this setup, if we query the model at a new feature vector $\vec x_0 \in \mathbb{R}^p$, then the prediction of decision tree $b$ is given by
$$
  \hat{y}_b(\vec x_0) = \sum_{i = 1}^{n} w_{b,0,i} \cdot y_i,
$$
where $w_{b,0,i} = \mathbb{I}\{\mathcal{T}_{b,0} = \mathcal{T}_{b,i}\} \cdot n_{b,i} / |\mathcal{T}_{b,0}|$ is the proportion of training data in $\mathcal{T}_{b,0}$ from observation $i$. The weights $w_{b,0,\cdot}$ can be interpreted as defining a local neighborhood around $\vec x_0$ for tree $b$ \citep{lin_random_2006}.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.7\linewidth]{variable_importance.pdf}
    \caption{\it Importance of predictor variables in random forest model. The response variable in this model is regressed NFL QBR, averaged across seasons in a player's career.}
    \label{fig:importance}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.7\linewidth]{3d_plot.pdf}
    \caption{\it Visualization of prediction function learned by random forest model. Passing yards per season and final-season strength of schedule are two of most important predictor variables in the model. Each point represents one player from the historical training set, and the color corresponds to the player's fitted value.}
    \label{fig:3d-plot}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.7\linewidth]{predicted_vs_actuals.pdf}
  \caption{\it Predicted and actual (regressed) QBR for all quarterbacks in the dataset. The vast majority (91\%) of actual QBRs are zero, meaning that the points along the x-axis represent 91\% of the data.}
  \label{fig:predicted-vs-actuals}
\end{figure}

@@

\begin{table}[H]
  \centering
  \input{top_ten.tex}
  \caption{\it Top ten career-average QBR predictions for quarterback prospects in the 2025 draft class. This includes draft-eligible players who played at an NCAA Division I FBS school during the 2024 season.}
  \label{tab:top-ten}
\end{table}

@@

These four quarterback prospects illustrate the model's interpretability and its alignment with real-world outcomes. Sanders and Ward were both initially expected to go in the first round of the 2025 draft. Sanders' draft position unexpectedly slid, and he ended up being taken in the fifth round in the draft. However, this is likely a result of his refusal to participate in the NFL Scouting Combine, coupled with a reported bad attitude in his interviews, rather than his college performance \citep{mckenna_what_2025}. In the end, Dart was the other quarterback taken in the first round of the draft.

\begin{table}[H]
  \resizebox{\textwidth}{!}{
    \input{side_by_side_similarity.tex}
  }
  \caption{\it Top ten similarity scores for each of the top four quarterback prospects in the 2025 NFL draft. The similarity score is the contribution made by each historical prospect to the weighted average which constitutes the reference prospect's prediction in the random forest model.}
  \label{tab:side-by-side-similarity}
\end{table}

It is worth noting that both Ward and Sanders started their college careers at teams that are not in the NCAA Division I FBS, as Ward played his first season at Incarnate Word and Sanders played his first two seasons at Jackson State. As a result, their college statistics used in the model do not include their statistics from these schools. We observe that Ward has the highest QBR prediction, which intuitively makes sense because he is widely regarded as the best quarterback in this year's draft class. Gabriel has a slightly higher QBR prediction than Sanders although Sanders was considered a better prospect. However, Gabriel finished higher than Sanders in Heisman voting, and Oregon had a much better record than Colorado. Dart's relatively low prediction compared to the other three makes sense because he wasn't considered a top quarterback prospect until his performance in the NFL Combine, which this model does not take into consideration. He also performed well in his interviews with teams \citep{traina_qb_2025}, which is not quantifiable.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.24\linewidth]{prospect_histogram_ward.pdf}
    \includegraphics[width=0.24\linewidth]{prospect_histogram_sanders.pdf}
    \includegraphics[width=0.24\linewidth]{prospect_histogram_gabriel.pdf}
    \includegraphics[width=0.24\linewidth]{prospect_histogram_dart.pdf} \\
    \includegraphics[width=0.24\linewidth]{prospect_similarity_ward.pdf}
    \includegraphics[width=0.24\linewidth]{prospect_similarity_sanders.pdf}
    \includegraphics[width=0.24\linewidth]{prospect_similarity_gabriel.pdf}
    \includegraphics[width=0.24\linewidth]{prospect_similarity_dart.pdf}
    \caption{\it Top: Histogram of training players' regressed QBR values weighted by similarity to the prospect being predicted. The vertical green line annotates each prospect's predicted QBR, which matches the mean of the weighted distribution. Bottom: Scatter plot showing individual similarity scores and regressed QBR for historical prospects.}
    \label{fig:prospect-plots}
\end{figure}

@@

\section{Acknolwedgments}

@@

\section{Author contributions}

EM downloaded the data, calculated the similarity scores, and drafted the manuscript. SP provided the research idea, converted EM's code into an R package, and revised the manuscript.

\section{Statements and Declarations}
\subsection{Ethical Considerations}

Not applicable

\subsection{Consent to Participate}

Not applicable

\subsection{Consent for Publication}

Not applicable

\subsection{Declaration of Conflicting Interest}

The authors declared no potential conflicts of interest with respect to the research, authorship, and/or publication of this article.

\subsection{Funding Statement}

The authors received no financial support for the research, authorship, and/or publication of this article.

\subsection{Code and Data Availability}

@@

\bibliographystyle{SageH}
\bibliography{random-forest-comps}
